C7: Fast software transactional memory for dynamic languages

### 2014-05-15 talk/dls2014/report/report.tex
move report to its own directory

Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading }and \emph{multiprocessing}.

@@

We focus on the \emph{threading }approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython and IronPython are implementations of
this. It requires great care in order to avoid deadlocks, which is why
we follow the TM approach that provides a \emph{direct }replacement
for the GIL. It does not require careful placing of locks in the right
spots. We will compare our work with Jython for evaluation.

@@

Our conflict detection works with \emph{object
  granularity}. Conceptually, it is based on \emph{read }and
\emph{write sets }of transactions.  Two transactions conflict if they
have accessed a common object that is now in the write set of at least
one of them.

The \emph{concurrency control }works partly \emph{optimistically} for
reading of objects, where conflicts caused by just reading an object
in transactions are detected only when the transaction that writes the
object actually commits. For write-write conflicts we are currently
\emph{pessimistic}: Only one transaction may have a certain object in
its write set at any point in time, others trying to write to it will
have to wait or abort.

We use \emph{lazy version management }to ensure that modifications by
a transaction are not visible to another transaction before the former
commits.

@@

As required for TM systems, we guarantee complete \emph{isolation
}and \emph{atomicity }for transactions at all times. Furthermore,
the isolation provides full \emph{opacity }to always guarantee a consistent
read set.

We support the notion of \emph{inevitable transactions }that are always
guaranteed to commit. There is always at most one such transaction
running in the system. We use this kind of transaction to provide
\emph{strong isolation} by running non-transactional code in the context
of inevitable transactions and to still provide the \emph{serializability}
of all transaction schedules.

@@

Generally speaking, the system is fully implemented in
software. However, we exploit some more advanced features of current
CPUs, especially \emph{memory segmentation, virtual memory, }and the
64-bit address space.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW) }on the
level of pages. Starting from the initial fully-shared configuration
(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

@@

Our TM system is designed as a library that covers all aspects around
transactions and object management. The library consists of two parts:
(I) It provides a simple interface to starting and committing
transactions, as well as the required read and write barriers. (II) It
also includes a \emph{garbage collector (GC) }that is closely
integrated with the TM part (e.g. it shares the write barrier). The
close integration helps in order to know more about the lifetime of an
object, as will be explained in the following sections.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment 0 is designated as
the \emph{sharing-segment }that is never assigned to a thread.

@@

The \textbf{first generation}, where objects are considered to be
\emph{young }and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of the nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

The \textbf{second generation}, where objects are considered to be
\emph{old }and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space. Furthermore,
we optimistically re-share pages that do not need to be private
anymore.

>>>>

Over the years, Python added multiple ways to provide concurrency and
parallelism to its applications. We want to highlight two of them,
namely \emph{threading} and \emph{multiprocessing}.

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from our interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython and IronPython are implementations of
this. It requires great care in order to avoid deadlocks, which is why
we follow the TM approach that provides a \emph{direct} replacement
for the GIL. It does not require careful placing of locks in the right
spots. We will compare our work with Jython for evaluation.

@@

Our conflict detection works with \emph{object
  granularity}. Conceptually, it is based on \emph{read} and
\emph{write sets} of transactions.  Two transactions conflict if they
have accessed a common object that is now in the write set of at least
one of them.

The \emph{concurrency control} works partly \emph{optimistically} for
reading of objects, where conflicts caused by just reading an object
in transactions are detected only when the transaction that writes the
object actually commits. For write-write conflicts we are currently
\emph{pessimistic}: Only one transaction may have a certain object in
its write set at any point in time, others trying to write to it will
have to wait or abort.

We use \emph{lazy version management} to ensure that modifications by
a transaction are not visible to another transaction before the former
commits.

@@

As required for TM systems, we guarantee complete \emph{isolation
}and \emph{atomicity} for transactions at all times. Furthermore,
the isolation provides full \emph{opacity} to always guarantee a consistent
read set.

We support the notion of \emph{inevitable transactions} that are always
guaranteed to commit. There is always at most one such transaction
running in the system. We use this kind of transaction to provide
\emph{strong isolation} by running non-transactional code in the context
of inevitable transactions and to still provide the \emph{serializability}
of all transaction schedules.

@@

Generally speaking, the system is fully implemented in
software. However, we exploit some more advanced features of current
CPUs, especially \emph{memory segmentation, virtual memory,} and the
64-bit address space.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages. Starting from the initial fully-shared configuration
(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

@@

Our TM system is designed as a library that covers all aspects around
transactions and object management. The library consists of two parts:
(I) It provides a simple interface to starting and committing
transactions, as well as the required read and write barriers. (II) It
also includes a \emph{garbage collector (GC)} that is closely
integrated with the TM part (e.g. it shares the write barrier). The
close integration helps in order to know more about the lifetime of an
object, as will be explained in the following sections.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment 0 is designated as
the \emph{sharing-segment} that is never assigned to a thread.

@@

The \textbf{first generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of the nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

The \textbf{second generation}, where objects are considered to be
\emph{old} and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space. Furthermore,
we optimistically re-share pages that do not need to be private
anymore.
